---
title: "Benchmarking against DESeq2 and edgeR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Benchmarking against DESeq2 and edgeR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

```{r, echo = FALSE}
suppressPackageStartupMessages({
    library(ggplot2)
    library(here)
})

# Plot benchmarks
plot_benchmark <- function(benchmark, title, design_formula) {
    the_plot <- ggplot(
        data = benchmark |> dplyr::filter(comparison != "disize"),
        mapping = aes(
            x = avg,
            fill = comparison
        )
    ) +
        geom_ribbon(aes(ymin = q5, ymax = q95), alpha = 0.1) +
        geom_ribbon(aes(ymin = q40, ymax = q60), alpha = 0.2) +
        geom_line(aes(x = avg, y = q50, color = comparison)) +
        scale_y_log10() +
        facet_wrap(
            ncol = 2,
            facets = c("sparsity", "mgt"),
            labeller = labeller(
                sparsity = function(x) paste0("Sparsity: ", x),
                mgt = function(x) paste0("Effect Magnitude: ", x)
            )
        ) +
        labs(
            title = title,
            subtitle = paste(
                "Design Formula:",
                paste(as.character(design_formula), collapse = " ")
            ),
            x = "Average Baseline Expression Level",
            y = "Relative Error",
            color = "Method",
            fill = "Method",
            linetype = "Method"
        ) +
        geom_hline(yintercept = 1.0, linetype = "dashed") +
        theme_classic()

    the_plot
}
```


# A Review of MoR and TMM Normalization

## Set Up

Let $y_{i,g}$ denote the observed count in a sample $i \in \mathcal{I}$ (with $n_i = |\mathcal{I}|$) for a gene $g \in \mathcal{G}$ (with $n_g = |\mathcal{G}|$) realized from a particular negative binomial distribution:

$$\begin{aligned}
    Y_{i,g} &\sim \text{NegBinom}(\mu_{i,g}, \phi_{g}) \\
    \mu_{i,g} &= \eta_{i,g} \cdot \rho_{i}
\end{aligned}$$

Where $\eta_{i,g}$ denotes the true magnitude of expression and $\rho_{i}$ denotes the "size factor" that scales this expression from its ground-truth.

## `DESeq2`'s Median of Ratios

We first focus on a particular gene $g$ and compare the observed count for a sample $i$ to the geometric average of counts across samples to get a ratio $R_{i,g}$:

$$
    R_{i,g} = \frac{ y_{i,g} }{( \prod_{i \in \mathcal{I}} y_{i,g} )^{ \frac{1}{n_i} } }
$$

We then take the median of these ratios to get our size factor estimates!

$$
    \hat{s}_{i} = \underset{g \in \mathcal{G}}{\text{median }} R_{i,g}
$$

## `edgeR`'s Trimmed Mean of M Values

We first "normalize" the observed count profile for a sample $i$ by the total number of counts $N_i = \sum_{g \in \mathcal{G}} y_{i,g}$ in order to get proportions:

$$
    y'_{i,g} = \frac{y_{i,g}}{N_i}
$$

We next select a reference sample $i^{\dagger} \in \mathcal{I}$ and compute both the ratio of log-transformed proportions and their weights:

$$\begin{aligned}
    R_{i,g} &= \frac{\log_2 y'_{i,g}}{\log_2 y'_{i^{\dagger}, g}} \\
    w_{i,g} &= \frac{ N_i - y_{i,g} }{ N_i y_{i,g} } + \frac{ N_{i^{\dagger}} - y_{i^{\dagger},g} }{ N_{i^{\dagger}} y_{i^{\dagger},g} }
\end{aligned}$$

We then filter the genes to a subset $\mathcal{G}'_i \subset \mathcal{G}$ by symmetrically "trimming" away the smallest and largest ratios for a sample $i$ to XX% of the original number (defaults to 70%).

We finally compute the "normalization" factor (distinct from a size factor) by taking the weighted average of these ratios and raising it to the second power:

$$
    \log_2 \hat{f}_i = \frac{ \sum_{g \in \mathcal{G}'_i} w_{i,g} R_{i,g} }{ \sum_{g \in \mathcal{G}'_i} w_{i,g} }
$$

The _size factor_ is then found by taking the product with the total number of observed counts:

$$
    \hat{s}_i = N_i \hat{f}_i
$$

> Note: This is referred to as the "effective library size" by the edgeR authors.

# Benchmarking Size Factor Estimation

We're interested in how well `DESeq2`, `edgeR` and `disize` estimate the true size factors underlying a given dataset.
The following simulation framework specifies: 

- A fixed experimental design.
- The amount of genes measured in the experiment.
- The fraction of genes who are **not** affected by a given measured covariate.
- The allowed magnitude of the covariate effect (*conditional on the gene being affected*).

For all experimental designs we construct various scenarios by varying the above parameters and then simulate a dataset corresponding to a given scenario:

```{r, eval=FALSE}
# Simulate a single dataset given its settings
simulate_dataset <- function(
    design_formula,
    metadata,
    n_genes,
    sparsity,
    mgt,
    avg) {
    # Simulate inverse overdispersion factors
    iodisps <- rlnorm(n_genes, log(10), 0.5)

    # Split design formula
    design <- split_formula(design_formula)

    # Construct model matrices...

    # Simulate size factors
    size_factors <- runif(length(levels(metadata$batch_id)), 0.1, 1.0) |>
        setNames(levels(metadata$batch_id))
    batch_effect <- size_factors[metadata$batch_id]

    # Simulate counts
    counts <- sapply(1:n_genes, function(g_i) {
        # Simulate fixed-effects
        sparse <- runif(ncol(fe_design)) < sparsity
        fe_coefs <- ifelse(sparse, 0.0, rnorm(ncol(fe_design), 0, mgt))

        # Simulate random-effects
        re_coefs <- lapply(lapply(remm$Ztlist, nrow), function(n_coefs) {
            sparse <- rep(runif(1) < sparsity, n_coefs)

            ifelse(sparse, 0.0, rnorm(n_coefs, sd = mgt))
        }) |>
            unlist()

        # Compute realized magnitude
        log_mu <- as.vector(
            rnorm(1, log(avg)) +
                fe_design %*% fe_coefs +
                re_design %*% re_coefs +
                log(batch_effect)
        )

        # Draw counts from negative binomial
        rnbinom(nrow(metadata), mu = exp(log_mu), size = iodisps[g_i])
    })

    # Cast to integers
    mode(counts) <- "integer"

    list(
        counts = counts,
        metadata = metadata,
        size_factors = log(
            size_factors / sum(size_factors) * length(size_factors)
        )
    )
}
```

## A Trivial Setting


```{r, echo = FALSE, fig.height=11}
benchmark <- read.table(
    here::here("benchmarks/size-factor-accuracy/data/scenario-1.tsv"),
    sep = "\t",
    header = TRUE
) |>
    dplyr::mutate(
        comparison = dplyr::case_match(
            comparison,
            "mor" ~ "MoR (DESeq/DESeq2)",
            "tmm" ~ "TMM (edgeR)",
            .default = comparison
        )
    )
plot_benchmark(benchmark, "A Trivial Setting", ~ (1 | donor_id))
```


## Comparing Two Conditions

```{r, echo = FALSE, fig.height=11}
benchmark <- read.table(
    here::here("benchmarks/size-factor-accuracy/data/scenario-2.tsv"),
    sep = "\t",
    header = TRUE
) |>
    dplyr::mutate(
        comparison = dplyr::case_match(
            comparison,
            "mor" ~ "MoR (DESeq/DESeq2)",
            "tmm" ~ "TMM (edgeR)",
            .default = comparison
        )
    )
plot_benchmark(
    benchmark,
    "Comparing Two Conditions",
    ~cond_id
)
```

## Multifactorial Design

```{r, echo = FALSE, fig.height=11}
benchmark <- read.table(
    here::here("benchmarks/size-factor-accuracy/data/scenario-2.tsv"),
    sep = "\t",
    header = TRUE
) |>
    dplyr::mutate(
        comparison = dplyr::case_match(
            comparison,
            "mor" ~ "MoR (DESeq/DESeq2)",
            "tmm" ~ "TMM (edgeR)",
            .default = comparison
        )
    )
plot_benchmark(
    benchmark,
    "Multifactorial Design",
    ~ cond_id:sex_id
)
```


# Impact on Downstream Differential Expression Analysis

Normalization is a small component of the transcriptomic workflow.
The impact on downstream analyses however is the main outcome of interest, as
size factors are only used to account for batch-effects.

To evaluate the impact of each normalization method on a typical downstream analysis, we 
specify a given experimental design and simulate a dataset similar to the benchmarks above:

```{r, eval=FALSE}
# Simulate a single dataset given its settings
simulate_dataset <- function(
    design_formula,
    metadata,
    n_genes,
    sparsity,
    mgt,
    avg) {
    # Simulate inverse overdispersion factors
    iodisps <- rlnorm(n_genes, log(10), 0.5)

    # Simulate size factors
    size_factors <- runif(length(levels(metadata$batch_id)), 0.1, 1.0) |>
        setNames(levels(metadata$batch_id))
    batch_effect <- size_factors[metadata$batch_id]

    # Construct sparsity mask
    mask <- rep(1.0, n_genes)
    mask[1:floor(sparsity * n_genes)] <- 0.0

    # Construct design matrix
    design <- model.matrix(design_formula, metadata)

    # Simulate counts
    counts <- sapply(1:n_genes, function(g_i) {
        # Simulate effects
        effects <- rnorm(ncol(design), sd = mgt)

        # Compute realized magnitude
        log_mu <- as.vector(
            rnorm(1, log(avg)) +
                design %*% effects * mask[g_i] +
                log(batch_effect)
        )

        # Draw counts from negative binomial
        rnbinom(nrow(design), mu = exp(log_mu), size = iodisps[g_i])
    })

    # Cast to integers
    mode(counts) <- "integer"

    list(
        counts = counts,
        metadata = metadata,
        size_factors = log(
            size_factors / sum(size_factors) * length(size_factors)
        ),
        null = mask == 0.0
    )
}
```

The only difference being we now keep track of which genes are truly null to 
benchmark the type 1 & type 2 error rates.

## Comparing Two Conditions

### Type 1 Error Rate
```{r, echo = FALSE, fig.height=11}
benchmark <- read.table(
    here::here("benchmarks/downstream-impact/data/scenario-1.tsv"),
    sep = "\t",
    header = TRUE
) |>
    dplyr::mutate(
        method = dplyr::case_match(
            method,
            "mor" ~ "MoR (DESeq/DESeq2)",
            "tmm" ~ "TMM (edgeR)",
            .default = method
        )
    ) |>
    dplyr::filter(
        method != "ground_truth"
    )

# Plot type 1 error
ggplot(
    data = benchmark,
    mapping = aes(x = avg, y = type_1_relative, color = method)
) +
    geom_line() +
    facet_wrap(
        ncol = 1,
        facets = c("sparsity", "mgt"),
        labeller = labeller(
            sparsity = function(x) paste0("Sparsity: ", x),
            mgt = function(x) paste0("Effect Magnitude: ", x)
        )
    ) +
    labs(
        title = "Comparing Two Conditions",
        subtitle = paste(
            "Design Formula:",
            paste(as.character(~cond_id), collapse = " ")
        ),
        x = "Average Baseline Expression Level",
        y = "Relative Type 1 Error",
        color = "Method"
    ) +
    geom_hline(yintercept = 1.0, linetype = "dashed") +
    theme_classic()
```

### Type 2 Error Rate
```{r, echo = FALSE, fig.height=11}
# Plot type 2 error
ggplot(
    data = benchmark,
    mapping = aes(x = avg, y = type_2_relative, color = method)
) +
    geom_line() +
    facet_wrap(
        ncol = 1,
        facets = c("sparsity", "mgt"),
        labeller = labeller(
            sparsity = function(x) paste0("Sparsity: ", x),
            mgt = function(x) paste0("Effect Magnitude: ", x)
        )
    ) +
    labs(
        title = "Comparing Two Conditions",
        subtitle = paste(
            "Design Formula:",
            paste(as.character(~cond_id), collapse = " ")
        ),
        x = "Average Baseline Expression Level",
        y = "Relative Type 2 Error",
        color = "Method"
    ) +
    geom_hline(yintercept = 1.0, linetype = "dashed") +
    theme_classic()
```